\LARGE \bf Learning Depth-Sensitive Conditional Random Fields for Semantic Segmentation of RGB-D Images

### 2013-09-12 paper.tex
Spell checker.
git-svn-id: https://svn.ais.uni-bonn.de/svn/staff/publications/icra_2014_mueller_nyu@4360 b2511e17-9ac0-488b-bc21-1d6d82e5baa2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We present a structured learning approach to semantic annotation of RGB-D
    images. 
    Our method learns to reason about spacial relations of objects and
    fuses low-level class predictions to a consistent interpretation of a
    scene. Our model incooporates color, depth and 3D scene features, on which
    an energy function is learned to directly optimize object class prediction using
    the loss-based maximum-margin principle of structural support vector machines.
    %With the proliferation of cheap % TODO urghs formulation
    %high-quality RGB-D sensors, it became possible to capture more and more
    %data not only as color images, but with additional depth information. The
    %depth channel provides important clues not only for scene structure, but
    %also for predicting semantic object classes.
    %We demonstrate how conditional random field learning, an approach
    %that has shown great promise for RGB images, can exploit depth information
    %to improve semantic interpretation of the scene.
    We evaluate our approach on the NYU V2 dataset of indoor scenes, a challenging
    dataset covering a wide variety of scene layouts and object classes. We
    hard-code much less information about the scene layout into our model then previous
    approaches, and instead learn object relations directly from the data. We
    find that our conditional random field approach improves upon previous
    work, setting a new state-of-the-art for the dataset.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{INTRODUCTION}
For robots to perform varied tasks in unstructured environments, understanding their
sourroundings is essential. We formulate the problem of semantic annotation of maps
as a dense labeling of RGB-D images into semantic classes. Dense labeling of objects
and structure classes allows for a detailed reasoning about the scene.

@@

We thereby extend the success of learned conditional random field models for
semantic segmentation in RGB images to the domain of 3D scenes.
Our emphasis lies on expoiting the additional depth and 3D information in all
procsessing steps, while relying on \emph{learning} to create a model that 
is adjusted to the properties of the sensor input and environment.

@@

We access the accuracy of our model on the challenging NYU Segmentation V2
dataset \citep{SilbermanECCV12}, where our model outperforms previous
approaches.  Our analysis shows that while our random forest model
already has competative performance, the superpixel-based grouping and in
pariculare the loss-based learning are integral ingredients of the success of our method.

@@

Work using RGB-D data only recently emmerged with
the availability of the NYU Depth Dataset V1~\citep{silberman2011indoor}.
\citet{silberman2011indoor} captured a wide variety of indoor scence using a structured light sensor,
and provided dense semantic object class annotation. Their work also introduced a baseline method
for semantic segmentation of RGB-D image, which used an approach based on superpixels, interest
point descriptors and a conditional random field. While pairwise potentials for the CRF were carfully
designed for the dataset, potentials were either directly set by hand or estimated using empirical
frequencies. This is in contrast to our work, which applies structured prediction techniques
to learn potentials automatically to optimize predictive performance.
\citet{ren2012rgb} focus on the design of features for semantic labeling of RGB-D data, and use
a hierarchical segmentation to provide context. While they also define a conditional random field on superpixels,
their model is again not learned, but a weighted potts model, using only a
probability of boundary map, and not taking spacial layout into account at all.

\citet{SilbermanECCV12} extendet the NYU Depth Dataset V1 to the NYU Depth Dataset V2 that we are also
using in this work.  Their focus is on inferring support relations in indoor
scence, such as objects resting on tables or shelfs, which in turn rest on the floor.
Their approach is based on robust estimation of 3D plane hypotheses, which are then jointly
optimized with support relations and structure classes.
\citet{SilbermanECCV12} use a complex pipeline, employing significant domain knowledge.
In this work, on the other hand, we try to learn all relevant domain specific features
directly from the data, which allows us to out-perform the work of \citet{SilbermanECCV12}
with respect to structure class segmentation.

\citet{couprie-iclr-13} approach the task of semantic segmentation of structure classes
in RGB-D using the paradigm of convolutional neural networks, extending
previous work of \citet{farabet-pami-13} and \citet{schulz2012learning}.
Similar to our approach, \citet{couprie-iclr-13} combine the output of a
pixel-based, low-level learning algorithm with an independent unsupervised
segmentation step. In contrast to their work, we improve our results by not
only averaging predictions within superpixels, but also explicitly learning
interactions between neighboring superpixels, favouring a consistent
interpretation of the whole image.

\citet{stueckler2013} proposed a real-time system for online learning and prediction of semantic classes.
Their main contribution is the online integration of low-level predictions over multiple frames
by estimating camera motion and building a 3D model of the scene.
We use the implementation of random forest provided by \citet{stueckler2013}, but instead of
integrating predictions over time, we focus on exploiting the structur within a single frame.

\section{LEARNING DEPTH-SENSITVE CONDITIONAL RANDOM FIELDS}
We take a conditional random field approach, whose nodes repesent a labeling of
superpixels.  We use an energy consisting of first and second order factors
(also called unary and pairwise potentials), with learned potential functions.
Let us denote the representation of an input image by $x$ and a labeling of
superpixels into semantic classes by $y$.  Then the general form of the energy is
\begin{equation}\eqlabel{energy}
    g(x, y) = \sum_{v \in V} \psi_v(x, y_v) + \sum_{(v, w) \in E} \psi_{v,w}(x, y_{v}, y_{w}).
\end{equation}
Here $V$ enumerates the superpixels, and $E\subset V \times V$ is a set of
edges, encoding neighborhood between superpixels.

We learn the unary and pairwise energy functions $\psi_v$ and $\psi_{v, w}$ from the training data
using a structural support vector machine. 
The concept of structural support vector machines allows for a principled,
maximum-margin based, loss-sensitive training of conditional random fields, that allows much
more complex interactions than the simple potts potentials that are often used in the literature.

In general, structural support vector machines learn the parameters $\theta$ of a predictor of the form
\begin{equation}\eqlabel{prediction}
f(x) = \argmax_{y \in \mathcal{Y}} \theta^T \Phi(x, y).
\end{equation}
We choose $\psi$ in \Eqref{energy} to be linear in the learnable parameters and
the data dependend features, resulting in a form equivalent to
\Eqref{prediction}. Our features are described in detail below.  We use the
$1$-lack formulation of the structural SVM~\citep{joachims2009cutting} and
solve the maximization in \Eqref{prediction} using a combination of fusion
moves~\citep{lempitsky2010fusion} and the AD$^3$ algorithm of
\citet{martins2011augmented}.  In contrast to graph-cut inference, fusion moves
and AD$^3$ can work with arbitrary potential function, and allow percise
learning using the structural support vector machine approach.

@@

\subsection{Unary Image Features}\seclabel{unaries}
Our method builds on the probability output of a random forest, trained for pixelwise
classification of the structure classes.
We use the GPU implementation provided by \citet{stueckler2013}.
The input for training are the full RGB-D images, transformed to Lab
color space. Each tree in the forest uses training pixel only from a subset of
training images. For each training image, an equal number of pixels for each
occurring class is sampled. Split features are given by difference of regions
on color or depth channels. Region size and offsets are normalized using the
depth at the target pixel.
We accumulate the probabilistic output for all pixels within a superpixel, and
use the resulting distribution as a feature for the unary node potentials in our CRF model.
We augment these prediction with another feature, based on the height of a superpixel in 3D.
This is a very informative feature, in particular to determin the floor.
To compute the height of a (super) pixel, we first find the ``up'' direction.
We use a very simple approach that we found effective: we cluster normal directions of all pixels
into 10 clusters using $k$-means, and use the one that is most parallel to the $Y$ direction, which
roughly corresponds to height in the dataset.
We then project the 3D pointcloud given by the depth along this direction, and normalize the result between
$0$ and $1$.
This procedure works robustly given there is any horizontal area in the image,
such as the ground or a table. A few scenes contain only walls and furniture,
and the approach fails for these. \Figref{height} illustrates a typical case and
one of the much rarer failure cases.
While we could use a more elaborat scheme,
such as the one from \citet{SilbermanECCV12}, we suspect that the feature is of
little use in scenes without horizontale surfaces.
\begin{figure}
    \begin{center}
        \includegraphics[width=\linewidth]{images/normal_feature}
    \end{center}
    \caption{%
        Visualization of one of the pairwise features, the similarity between superpixel normals.
        The image shows the zoom-in of a bedroom scene, together with the superpixel oversegmentation.
        Lines connect adjacent superpixels, and line-stregth gives the magnitude of the orientation similarity.
    \figlabel{normal_feature}}
\end{figure}

@@

A visualization of the impact of each processing step can be found in
\Figref{comparison}, which shows prediction results on the test set.  The for
prediction methods correspond to the rows of \Tabref{results}.  The difference
between the SVM and CRF approaches are clearly visible, with the CRF producing
results that are very close to the ground truth in serveral complex scenes.

We found that our approach improves results most for scenes with a clear geometric structure,
which is not surprising. We see that evidence from the random forest is often very noisy, and biased
away from the ``props'' class. While the unstructured SVM can correct somewhat
for the class imbalance, it has no way to make larger areas consistent, which
the conditional random field can.
On the other hand, performance of the CRF deteriorates sligthly on very crowded
scenes with a mixture of small furniture and prop objects, as
can be seen in the two right-most images. In these scenes, depth information is
often noisy, and it is hard to make geometric statements on the superpixel
level. As the input from the random forest is also often of low quality for
crowded scenes, the conditional random field has little chance to recover.

@@

\Figref{weights} visualizes some learned potential functions.
Higher values correspond to favored configurations. One can see that the
vertical alignment potential expresses that the floor is much more likely to be below
other classes. It also encodes the fact that props
rest on furniture, but not the other way around.
The potential of the depth feature encodes, for example, that the ground
is usually behind the other classes, while furniture is in front of structures,
such as the wall.
Interestingly, the potential functions are not anti-symmetric, and forcing them to be so
degrates the results. This suggests that the direction of connecting edges, going from
the top left to the bottom right in the image, is also exploited by the potentials.

@@

While our approach allows modelling of spacial relations, these are limited to
local interactions. In future work, these interactions could be extended to
larger areas using latent variable models~\citep{dann2012pottics} or higher
order potentials~\citep{kohli2009robust}.
Another possible line of future work is to combine our approach with a more
task-specific one, directly including support plane assumptions into the model, as done by
\citet{SilbermanECCV12}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
    We present a structured learning approach to semantic annotation of RGB-D
    images. 
    Our method learns to reason about spatial relations of objects and
    fuses low-level class predictions to a consistent interpretation of a
    scene. Our model incorporates color, depth and 3D scene features, on which
    an energy function is learned to directly optimize object class prediction using
    the loss-based maximum-margin principle of structural support vector machines.
    %With the proliferation of cheap % TODO urghs formulation
    %high-quality RGB-D sensors, it became possible to capture more and more
    %data not only as color images, but with additional depth information. The
    %depth channel provides important clues not only for scene structure, but
    %also for predicting semantic object classes.
    %We demonstrate how conditional random field learning, an approach
    %that has shown great promise for RGB images, can exploit depth information
    %to improve semantic interpretation of the scene.
    We evaluate our approach on the NYU V2 dataset of indoor scenes, a challenging
    dataset covering a wide variety of scene layouts and object classes. We
    hard-code much less information about the scene layout into our model then previous
    approaches, and instead learn object relations directly from the data. We
    find that our conditional random field approach improves upon previous
    work, setting a new state-of-the-art for the dataset.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{INTRODUCTION}
For robots to perform varied tasks in unstructured environments, understanding their
surroundings is essential. We formulate the problem of semantic annotation of maps
as a dense labeling of RGB-D images into semantic classes. Dense labeling of objects
and structure classes allows for a detailed reasoning about the scene.

@@

We thereby extend the success of learned conditional random field models for
semantic segmentation in RGB images to the domain of 3D scenes.
Our emphasis lies on exploiting the additional depth and 3D information in all
processing steps, while relying on \emph{learning} to create a model that 
is adjusted to the properties of the sensor input and environment.

@@

We access the accuracy of our model on the challenging NYU Segmentation V2
dataset \citep{SilbermanECCV12}, where our model outperforms previous
approaches.  Our analysis shows that while our random forest model
already has competitive performance, the superpixel-based grouping and in
particular the loss-based learning are integral ingredients of the success of our method.

@@

Work using RGB-D data only recently emerged with
the availability of the NYU Depth Dataset V1~\citep{silberman2011indoor}.
\citet{silberman2011indoor} captured a wide variety of indoor scene using a structured light sensor,
and provided dense semantic object class annotation. Their work also introduced a baseline method
for semantic segmentation of RGB-D image, which used an approach based on superpixels, interest
point descriptors and a conditional random field. While pairwise potentials for the CRF were carefully
designed for the dataset, potentials were either directly set by hand or estimated using empirical
frequencies. This is in contrast to our work, which applies structured prediction techniques
to learn potentials automatically to optimize predictive performance.
\citet{ren2012rgb} focus on the design of features for semantic labeling of RGB-D data, and use
a hierarchical segmentation to provide context. While they also define a conditional random field on superpixels,
their model is again not learned, but a weighted Potts model, using only a
probability of boundary map, and not taking spacial layout into account at all.

\citet{SilbermanECCV12} extended the NYU Depth Dataset V1 to the NYU Depth Dataset V2 that we are also
using in this work.  Their focus is on inferring support relations in indoor
scenes, such as objects resting on tables or shelfs, which in turn rest on the floor.
Their approach is based on robust estimation of 3D plane hypotheses, which are then jointly
optimized with support relations and structure classes.
\citet{SilbermanECCV12} use a complex pipeline, employing significant domain knowledge.
In this work, on the other hand, we try to learn all relevant domain specific features
directly from the data, which allows us to out-perform the work of \citet{SilbermanECCV12}
with respect to structure class segmentation.

\citet{couprie-iclr-13} approach the task of semantic segmentation of structure classes
in RGB-D using the paradigm of convolutional neural networks, extending
previous work of \citet{farabet-pami-13} and \citet{schulz2012learning}.
Similar to our approach, \citet{couprie-iclr-13} combine the output of a
pixel-based, low-level learning algorithm with an independent unsupervised
segmentation step. In contrast to their work, we improve our results by not
only averaging predictions within superpixels, but also explicitly learning
interactions between neighboring superpixels, favoring a consistent
interpretation of the whole image.

\citet{stueckler2013} proposed a real-time system for online learning and prediction of semantic classes.
Their main contribution is the online integration of low-level predictions over multiple frames
by estimating camera motion and building a 3D model of the scene.
We use the implementation of random forest provided by \citet{stueckler2013}, but instead of
integrating predictions over time, we focus on exploiting the structure within a single frame.

\section{LEARNING DEPTH-SENSITVE CONDITIONAL RANDOM FIELDS}
We take a conditional random field approach, whose nodes represent a labeling of
superpixels.  We use an energy consisting of first and second order factors
(also called unary and pairwise potentials), with learned potential functions.
Let us denote the representation of an input image by $x$ and a labeling of
superpixels into semantic classes by $y$.  Then the general form of the energy is
\begin{equation}\eqlabel{energy}
    g(x, y) = \sum_{v \in V} \psi_v(x, y_v) + \sum_{(v, w) \in E} \psi_{v,w}(x, y_{v}, y_{w}).
\end{equation}
Here $V$ enumerates the superpixels, and $E\subset V \times V$ is a set of
edges, encoding neighborhood between superpixels.

We learn the unary and pairwise energy functions $\psi_v$ and $\psi_{v, w}$ from the training data
using a structural support vector machine. 
The concept of structural support vector machines allows for a principled,
maximum-margin based, loss-sensitive training of conditional random fields, that allows much
more complex interactions than the simple Potts potentials that are often used in the literature.

In general, structural support vector machines learn the parameters $\theta$ of a predictor of the form
\begin{equation}\eqlabel{prediction}
f(x) = \argmax_{y \in \mathcal{Y}} \theta^T \Phi(x, y).
\end{equation}
We choose $\psi$ in \Eqref{energy} to be linear in the learnable parameters and
the data-depended features, resulting in a form equivalent to
\Eqref{prediction}. Our features are described in detail below.  We use the
$1$-lack formulation of the structural SVM~\citep{joachims2009cutting} and
solve the maximization in \Eqref{prediction} using a combination of fusion
moves~\citep{lempitsky2010fusion} and the AD$^3$ algorithm of
\citet{martins2011augmented}.  In contrast to graph-cut inference, fusion moves
and AD$^3$ can work with arbitrary potential function, and allow precise
learning using the structural support vector machine approach.

@@

\subsection{Unary Image Features}\seclabel{unaries}
Our method builds on the probability output of a random forest, trained for pixel-wise
classification of the structure classes.
We use the GPU implementation provided by \citet{stueckler2013}.
The input for training are the full RGB-D images, transformed to Lab
color space. Each tree in the forest uses training pixel only from a subset of
training images. For each training image, an equal number of pixels for each
occurring class is sampled. Split features are given by difference of regions
on color or depth channels. Region size and offsets are normalized using the
depth at the target pixel.
We accumulate the probabilistic output for all pixels within a superpixel, and
use the resulting distribution as a feature for the unary node potentials in our CRF model.
We augment these prediction with another feature, based on the height of a superpixel in 3D.
This is a very informative feature, in particular to determine the floor.
To compute the height of a (super) pixel, we first find the ``up'' direction.
We use a very simple approach that we found effective: we cluster normal directions of all pixels
into 10 clusters using $k$-means, and use the one that is most parallel to the $Y$ direction, which
roughly corresponds to height in the dataset.
We then project the 3D point cloud given by the depth along this direction, and normalize the result between
$0$ and $1$.
This procedure works robustly given there is any horizontal area in the image,
such as the ground or a table. A few scenes contain only walls and furniture,
and the approach fails for these. \Figref{height} illustrates a typical case and
one of the much rarer failure cases.
While we could use a more elaborate scheme,
such as the one from \citet{SilbermanECCV12}, we suspect that the feature is of
little use in scenes without horizontal surfaces.
\begin{figure}
    \begin{center}
        \includegraphics[width=\linewidth]{images/normal_feature}
    \end{center}
    \caption{%
        Visualization of one of the pairwise features, the similarity between superpixel normals.
        The image shows the zoom-in of a bedroom scene, together with the superpixel oversegmentation.
        Lines connect adjacent superpixels, and line-strength gives the magnitude of the orientation similarity.
    \figlabel{normal_feature}}
\end{figure}

@@

A visualization of the impact of each processing step can be found in
\Figref{comparison}, which shows prediction results on the test set.  The for
prediction methods correspond to the rows of \Tabref{results}.  The difference
between the SVM and CRF approaches are clearly visible, with the CRF producing
results that are very close to the ground truth in several complex scenes.

We found that our approach improves results most for scenes with a clear geometric structure,
which is not surprising. We see that evidence from the random forest is often very noisy, and biased
away from the ``props'' class. While the unstructured SVM can correct somewhat
for the class imbalance, it has no way to make larger areas consistent, which
the conditional random field can.
On the other hand, performance of the CRF deteriorates slightly on very crowded
scenes with a mixture of small furniture and prop objects, as
can be seen in the two right-most images. In these scenes, depth information is
often noisy, and it is hard to make geometric statements on the superpixel
level. As the input from the random forest is also often of low quality for
crowded scenes, the conditional random field has little chance to recover.

@@

\Figref{weights} visualizes some learned potential functions.
Higher values correspond to favored configurations. One can see that the
vertical alignment potential expresses that the floor is much more likely to be below
other classes. It also encodes the fact that props
rest on furniture, but not the other way around.
The potential of the depth feature encodes, for example, that the ground
is usually behind the other classes, while furniture is in front of structures,
such as the wall.
Interestingly, the potential functions are not anti-symmetric, and forcing them to be so
degrades the results. This suggests that the direction of connecting edges, going from
the top left to the bottom right in the image, is also exploited by the potentials.

@@

While our approach allows modeling of spacial relations, these are limited to
local interactions. In future work, these interactions could be extended to
larger areas using latent variable models~\citep{dann2012pottics} or higher
order potentials~\citep{kohli2009robust}.
Another possible line of future work is to combine our approach with a more
task-specific one, directly including support plane assumptions into the model, as done by
\citet{SilbermanECCV12}.
